% $Id: abstract.tex 1789 2010-09-28 16:30:23Z jabriffa $

\chapter*{Abstract}

This project investigates the practical cost of running symmetric cryptography
inside a zero-knowledge virtual machine. The implementation target is RISC Zero,
and the benchmark set covers three algorithms with different structures:
AES-128, Salsa20, and LowMC. A configurable benchmark harness was developed to
run each target, collect machine-readable JSON output, aggregate results, and
produce comparison plots.

The current draft's primary technical path is a LowMC optimization case study,
chosen because baseline benchmarks showed it was the dominant proving bottleneck
in the current workload set. This scope is not fully locked yet: the report
tracks a pending decision between (i) deeper LowMC variant comparison and
(ii) implementing one complete authenticated-encryption algorithm.

For the LowMC path, the optimized design combines three changes: a vectorized
SWAR S-box layer, matrix hot-path restructuring, and reduced guest-side setup
through precomputed round-key inputs. Validation includes roundtrip checks,
reference ciphertext checks, and dedicated tests for the vectorized S-box logic.

In the current draft benchmark snapshot, LowMC proving time decreases from
695.05 s to 517.21 s, with user cycles reduced from 10,388,894 to 7,919,919.
These values are provisional and may be replaced by the final campaign aligned
to the locked scope. By contrast, the AES optimization workspace currently acts
as scaffolding for S-box experimentation rather than a completed performance
improvement, and no multi-zkVM comparison is claimed.

Overall, the study shows that proving cost in this setting is strongly shaped by
algorithm structure and by low-level implementation choices in nonlinear and
matrix-heavy paths. The report also documents limitations in statistical rigor
and scope, and outlines future work for stronger comparative and reproducible
evaluation.
