\chapter{LowMC Optimization Design and Implementation}

\textbf{Draft scope note: this chapter reflects the current LowMC-focused track.
If the final scope pivots to a full authenticated-encryption implementation,
this chapter will be reduced to supporting baseline material.}

\section{Motivation for Optimizing LowMC}

In the current draft benchmark snapshot, LowMC has the highest proving cost among the
tested cipher targets by a large margin. Baseline LowMC proving time is
695.05 seconds, compared with approximately 14.84 seconds for AES and
29.79 seconds for Salsa20 under their configured workloads.

This currently makes LowMC the most impactful optimization target in the
available timescale.

\section{Vectorized SWAR S-box Layer}

\subsection{Baseline Behavior}

The baseline implementation applies each 3-bit S-box lane serially using table
lookups across 49 lanes per round. This is straightforward, but it performs
repeated per-lane bit extraction and reinsertion.

\subsection{Optimized Design}

The optimized implementation introduces a SWAR-style vectorized substitution
layer in \texttt{lowmc-r0-optimised/lowmc-core/src/lib.rs}. Three precomputed masks
select lane bit-planes (positions $3i$, $3i+1$, and $3i+2$). After alignment
shifts, boolean equations are evaluated across all lanes in parallel, and
results are shifted back and merged.

Forward equations in aligned lane form are:

\begin{equation}
y_0 = x_0 \oplus x_1 \oplus x_2 \oplus (x_1 \wedge x_2)
\end{equation}

\begin{equation}
y_1 = x_1 \oplus x_2 \oplus (x_0 \wedge x_2)
\end{equation}

\begin{equation}
y_2 = x_2 \oplus (x_0 \wedge x_1)
\end{equation}

Only the first 147 bits (49 lanes $\times$ 3 bits) are replaced; the linear tail
is preserved.

\subsection{Expected Effect in zkVM Context}

This design reduces per-lane control overhead and favors branchless,
boolean-heavy computation. Given the operation-level directional benchmarks in
the repository, this is a plausible strategy for lowering user-cycle pressure in
this path.

\section{Matrix Hot-path Optimization}

Matrix multiplication is repeated heavily in LowMC rounds. The baseline path used
row-wise parity from repeated \texttt{count\_ones()} operations with per-bit output
setter calls.

The optimized path applies two changes:

\begin{itemize}
\item branchless XOR-fold parity (\texttt{parity\_u64}) instead of popcount-based
parity in key-matrix multiplication;
\item transposed matrix-column representation with direct XOR accumulation in
\texttt{multiply\_block\_matrix\_columns}, reducing repeated bit setter overhead in
the block-matrix path.
\end{itemize}

These changes target repeated matrix work where small per-iteration savings can
accumulate over rounds.

\section{Reduced Guest Setup and Precomputed Round Keys}

In the baseline design, the guest receives key matrices and computes round keys
internally. In the optimized design, host code precomputes round keys and passes
them directly.

Concretely:

\begin{itemize}
\item baseline guest spec includes key matrices and key bytes;
\item optimized guest spec includes precomputed \texttt{round\_keys} and omits
guest-side round-key derivation.
\end{itemize}

This reduces guest work and input-processing overhead, which is relevant for
proof cost even when algorithmic behavior is unchanged.

\section{Validation and Correctness Assurance}

Optimization correctness is validated through several tests in
\texttt{lowmc-r0-optimised/lowmc-core/src/lib.rs}, including:

\begin{itemize}
\item \texttt{vectorized\_sbox\_matches\_reference\_tables};
\item \texttt{roundtrip\_encrypt\_decrypt};
\item \texttt{matches\_reference\_cpp\_ciphertext};
\item \texttt{precomputed\_columns\_roundtrip\_matches\_encrypt\_decrypt}.
\end{itemize}

Together, these checks cover forward/inverse substitution behavior, roundtrip
integrity, reference ciphertext matching, and precomputed-data reconstruction.

\textbf{TODO: append representative test logs in an appendix for stronger
traceability.}

\section{Implementation Trade-offs}

The optimization reduces proving cost but increases implementation complexity.
Bit-lane masking, cross-word shifts, and transposed-matrix logic are less
readable than serial per-lane code. This raises maintenance and debugging cost,
and makes robust tests mandatory for future changes.

For the LowMC track, this trade-off is acceptable because measurable
proof-cost reductions are the primary objective.

\textbf{TODO: add a concise before/after pseudocode figure for the S-box layer
to improve readability for non-specialist examiners.}
