\chapter{Background and Literature Review}

\section{Verifiable Computation and zkVMs}

Modern verifiable-computation systems aim to prove that a computation was
executed correctly without requiring every verifier to re-execute it.
Pinocchio showed an influential route to practical verifiable computation, but
still required specialized circuit-oriented workflows
\cite{parno2013pinocchio}. Later transparent proof systems, including STARK
constructions, reduced trusted-setup concerns and improved scalability in
practice \cite{bensasson2018stark}.

From an engineering perspective, manual circuit design remains a high-friction
workflow for software teams. zkVMs address this by running conventional
instruction sets while generating proofs of execution. The trade-off is that
developer convenience comes with overhead from virtualization, memory handling,
and proof-system integration.

RISC Zero is a reasonable target for this project because it provides:

\begin{itemize}
\item a clear host/guest execution model;
\item a Rust-based toolchain that aligns with the repository implementation;
\item direct access to execution statistics (cycles and segments) needed for
benchmark interpretation \cite{risczeroDocs};
\item emerging evidence of broader software-provenance applications
\cite{ron2026provenance}.
\end{itemize}

This does not imply RISC Zero is universally superior to other zkVMs. It means
it is suitable for a focused, reproducible evaluation under final-year project
constraints.

\subsection{How RISC Zero Works in This Project (High-Level)}

At a high level, RISC Zero lets a prover convince a verifier that a specific
program ran correctly on specific inputs, without the verifier replaying the
full computation \cite{risczeroDocs}. In this repository, that flow is:

\begin{enumerate}
\item \textbf{Compile guest code.} Cryptographic workload logic is compiled into
RISC-V guest binaries that run inside the zkVM.
\item \textbf{Prepare host inputs.} Host code serializes benchmark inputs
(plaintext, keys, parameters, expected checks) and launches guest execution.
\item \textbf{Prove execution.} The prover executes the guest, records the
execution trace, and generates a receipt proving the trace is valid for that
guest program.
\item \textbf{Expose public outputs.} The guest commits selected outputs to the
journal (for example, correctness flags or benchmark result payloads).
\item \textbf{Verify cheaply.} The host (or any verifier) checks the receipt
against the expected guest program identity and journal contents, without
re-running the full workload.
\end{enumerate}

For this dissertation, the practical value is twofold: (i) benchmark outputs are
tied to a verifiable execution record, and (ii) the proving process exposes cost
signals (for example cycle and segment statistics) that can be analyzed across
cipher implementations \cite{risczeroDocs}.

\section{Symmetric Ciphers Under zk Cost Models}

This project compares three structures:

\begin{itemize}
\item AES-128 as a standard block cipher with byte-level substitution and
linear diffusion;
\item LowMC as a design that minimizes nonlinear operations to suit MPC/FHE-like
cost models \cite{albrecht2015ciphers};
\item Salsa20 as an ARX stream cipher with no S-box table lookups
\cite{bernstein2008salsa20}.
\end{itemize}

The critical question is not ``which cipher is globally faster'' but ``which
structural properties dominate proving overhead in this zkVM setup.'' In
practice, this depends on how algorithm structure maps onto guest code,
serialization overhead, and repeated matrix or substitution operations.

An important observation from this project's measurements is that theoretical
design intent (for example, LowMC's reduced nonlinearity motivation) does not
automatically guarantee low proving cost in a concrete zkVM implementation.
Representation choices and data movement still matter.

\section{S-box Design and Optimization Trade-offs}

S-box implementation strategy is central to this dissertation because it drives
the main optimization chapter. Existing literature shows multiple valid
realizations: lookup tables, logic-minimized combinational networks, and
depth-oriented circuit variants
\cite{canright2005compactsbox,boyar2010logicmin,boyar2011depth16}.

For normal CPU software, lookup-table implementations can be simple and fast,
but they may expose timing/cache side-channel concerns under some deployment
conditions \cite{osvik2006cache}. In contrast, boolean-heavy or bitsliced
approaches can improve constant-time behavior and better align with gate or
constraint-oriented reasoning.

Inside zkVM execution, these trade-offs shift again. The cost signal is not only
instruction latency but also proof-system behavior for control flow,
memory-access patterns, and nonlinear logic realization. This motivates the
project decision to favor branchless boolean-heavy experimentation in hot
LowMC paths, supported by operation-level directional measurements and
correctness tests.

\section{Trust, Implementation Attacks, and ASAs}

Algorithm substitution attacks (ASAs) show that a cryptographic implementation
can be maliciously altered while preserving apparently correct external behavior
\cite{bellare2014asa}. Separately, Thompson's ``trusting trust'' argument
demonstrates that build chains and compilers can undermine source-level trust
\cite{thompson1984trustingtrust}.

zkVM-based provenance approaches can strengthen some trust properties by proving
what code executed, with what inputs, under which execution trace
\cite{ron2026provenance}. However, this does not ``solve'' implementation trust
by itself. Trust still depends on source review, build process integrity,
benchmark fairness, and correct interpretation of results.

Accordingly, this report treats ASAs and trust as motivation plus critical
discussion, not as a completed attack-detection experiment.

\section{Literature Gap Addressed by This Project}

The reviewed literature provides strong foundations on proof systems, cipher
design, and implementation security, but it does not answer this report's core
engineering question directly: how expensive are practical symmetric-cipher
workloads inside a specific zkVM implementation, and what optimization changes
produce measurable wins in that context?

This dissertation currently addresses that gap with repository-grounded baseline
evidence: multi-target RISC Zero benchmarks and a reproducible harness.

The final implementation emphasis is still scope-dependent. The current draft
path centers on LowMC optimization comparisons, but a full
authenticated-encryption implementation remains an alternative final track (see
Chapter 3).

\textbf{TODO: add one or two recent zkVM benchmarking papers (2024--2026) after
bibliographic verification, then expand this chapter's comparison subsection.}
