\chapter{System Design and Benchmarking Methodology}

\section{Repository and Benchmark Architecture}

The repository is organized as a monorepo where each benchmark target is
isolated in its own directory, for example \texttt{aes-r0/},
\texttt{aes-r0-optimised/}, \texttt{salsa-r0/}, \texttt{lowmc-r0/},
\texttt{lowmc-r0-optimised/}, and \texttt{operation-bnchmrk-r0/}. This structure
supports independent iteration per target while preserving a common benchmark
harness interface.

The harness itself is configuration-driven:

\begin{itemize}
\item \texttt{runner.py} executes enabled targets and captures JSON output,
stdout/stderr logs, and per-trial status.
\item \texttt{aggregate.py} computes summary statistics per benchmark ID.
\item \texttt{plot.py} generates visualizations from aggregated results.
\item \texttt{schema.md} defines the expected JSON contract.
\end{itemize}

For each run, outputs are written under
\texttt{artifacts/benchmarks/<run-id>/} with a manifest, raw records,
aggregated summary, and plots.

\section{Benchmark Metrics}

The primary metrics reported are listed in
Table \ref{tab:metric-definitions}.

\begin{table}[t]
  \begin{center}
    \begin{tabular}{|p{0.26\textwidth}|p{0.64\textwidth}|}
      \hline
      \textbf{Metric} & \textbf{Definition and interpretation} \\
      \hline
      \texttt{prove\_seconds} & Wall-clock time to generate the zk proof for a run. This is the main cost metric in this project. \\
      \hline
      \texttt{verify\_seconds} & Wall-clock time to verify the generated receipt. Usually much smaller than proving time in reported runs. \\
      \hline
      \texttt{total\_seconds} & \texttt{prove\_seconds + verify\_seconds}. \\
      \hline
      \texttt{total\_cycles} & Total cycle count reported by prover statistics. Includes user, paging, and reserved components. \\
      \hline
      \texttt{user\_cycles} & Cycles attributed to guest computation; used as a directional implementation-cost proxy. \\
      \hline
      \texttt{paging\_cycles} & Cycles attributed to memory paging overhead in the zkVM runtime. \\
      \hline
      \texttt{reserved\_cycles} & Runtime/prover-reserved cycle component; useful for identifying non-user overhead shifts. \\
      \hline
    \end{tabular}
  \end{center}
  \caption{Benchmark metric definitions used in this report}
  \label{tab:metric-definitions}
\end{table}

For aggregated statistics, each metric includes count, mean, median, p95,
standard deviation, min, and max.

\section{Experimental Setup}

The current draft snapshot used for reporting examples is run
\texttt{20260310-010549Z}. Its manifest records:

\begin{itemize}
\item system: Darwin 25.3.0;
\item architecture: arm64 (processor reported as arm);
\item python runtime: 3.9.6;
\item default harness settings: \texttt{trials = 1},
\texttt{timeout\_sec = 1000};
\item target commands: release-profile \texttt{cargo run -p host --release -- --json}
(operation benchmark adds \texttt{--iterations 100}).
\end{itemize}

\textbf{TODO: add full machine specification (CPU model, core count, RAM,
thermal/power mode) used for the final scope-aligned benchmark runs.}

\textbf{TODO: add explicit rationale for chosen payload sizes and iteration
counts (AES 64-byte payload, Salsa20 1024-byte payload, operation benchmark
100 iterations).}

\section{Fairness and Reproducibility Considerations}

To improve reproducibility:

\begin{itemize}
\item all target commands are versioned in \texttt{bench-harness/config.toml};
\item each run records platform and command metadata in
\texttt{run\_manifest.json};
\item raw logs and per-trial records are preserved;
\item aggregation is script-based, avoiding manual transcription.
\end{itemize}

Fairness is handled by keeping one benchmark harness and common execution mode
(zk, release profile) across targets. However, strict equivalence is limited by
algorithm-specific workloads and by target-specific input shapes.

\section{Threats to Methodological Validity}

TODO: Fill out this section
