\chapter{Introduction}

\section{Problem Context}

Symmetric cryptography is routinely used in real systems because it is fast and
well studied, but this performance intuition comes from conventional execution
environments. In a zero-knowledge virtual machine (zkVM), the relevant cost is
not only runtime but also proof-generation overhead. Operations that appear
cheap on standard CPUs can become expensive once execution must be encoded and
proved.

This project studies that gap by benchmarking representative symmetric
algorithms inside the RISC Zero zkVM and then optimizing the most costly case.
The central question is practical rather than purely theoretical: which
algorithmic and implementation choices materially change proving cost in a real
zkVM workflow?

\section{Why the Problem Matters}

Verifiable execution is increasingly relevant for security-sensitive software,
auditable cryptographic workflows, and reproducible computation. Prior work in
verifiable computation establishes strong foundations
\cite{parno2013pinocchio,bensasson2018stark}, and recent systems work shows how
zkVM-based compilation can improve software provenance and auditability
\cite{ron2026provenance}. However, practical adoption depends on measurable
cost, especially for cryptographic code paths that are often non-linear and
bit-manipulation heavy.

If proving overhead is too high, deployment claims become weak regardless of
correctness guarantees. For this reason, this dissertation emphasizes measured
costs, reproducibility, and controlled claims.

\section{Why the Problem Is Hard}

Three constraints make this problem difficult.

First, zk proving costs are sensitive to structure. Cipher components such as
S-boxes, matrix operations, and memory access patterns interact with the zkVM
execution model in ways that are not captured by native timing alone.

Second, fair benchmarking is non-trivial. Different algorithms naturally use
different workloads and parameter sets, so naive comparison can overstate or
misstate performance differences.

Third, optimization work can easily produce code that is faster but harder to
reason about. This raises validation, maintenance, and trust concerns that must
be documented explicitly.

\section{Project Approach}

The implemented approach is staged:

\begin{enumerate}
\item Build a reusable benchmarking harness for RISC Zero targets, with
structured JSON outputs and plotting support.
\item Implement and benchmark AES-128, Salsa20, and LowMC in RISC Zero.
\item After scope lock, complete one primary track:
\begin{itemize}
\item LowMC variant comparison (S-box-optimised vs more circuit-oriented), or
\item one full authenticated-encryption implementation and benchmark path.
\end{itemize}
\end{enumerate}


\section{Main Contributions}

The dissertation currently targets the following draft contributions:

\begin{itemize}
\item a configurable benchmark harness for zk runs, aggregation, and plotting;
\item a set of benchmarkable symmetric targets in RISC Zero;
\item a scope-dependent implementation contribution (LowMC variant study or full authenticated-encryption implementation);
\item an evidence-based evaluation of proof-cost trade-offs across algorithm and
implementation structure.
\end{itemize}

\section{Headline Results}

TODO: This section will include a overview of the results found using benchmarking

\section{Why RISC Zero Was Chosen}

RISC Zero was selected as the implementation target because it offers a mature
Rust developer workflow, straightforward host/guest separation, proof APIs that
expose cycle statistics needed for analysis, and strong public documentation
\cite{risczeroDocs}. This made it a practical platform for a final-year project
focused on measurable engineering outcomes rather than proof-system design.

Other options were considered but not taken forward in this draft. A second
zkVM target (for example SP1) would improve platform-level comparison, but it
would also require duplicated integration work (guest/host plumbing,
benchmark-command standardization, and validation) within limited project time.
Circuit-first approaches were also not selected as the primary path because this
project's objective is implementation-level benchmarking with minimal redesign of
algorithm code. Under these constraints, RISC Zero provided the best balance of
engineering velocity, observability, and reproducibility for the current phase.

\section{Scope Boundaries and Limitations}

The current project scope is intentionally narrower than the original proposal,
but the final track is still being locked (see Chapter 3). At draft stage,
RISC Zero is the only experimentally evaluated zkVM, and ASA discussion remains
motivation/critical context rather than formal attack evaluation.

Whether authenticated-encryption modes are implemented in this report depends on
the final track decision: they are out of the current LowMC-focused baseline,
but remain a possible scope pivot.

The benchmark campaign used for primary reporting also has statistical
limitations: some reported runs have one trial only, and workloads differ across
algorithms. These limitations are discussed explicitly in Chapter 7.

\section{Report Structure}

The rest of this dissertation is organized as follows:

\begin{itemize}
\item Chapter 2 reviews verifiable computation, zkVM context, cipher structure,
and S-box design literature.
\item Chapter 3 defines working scope, objectives, success criteria, and project
timeline pivots.
\item Chapter 4 presents the system architecture and benchmarking methodology.
\item Chapter 5 documents implemented cipher targets in RISC Zero.
\item Chapter 6 details the LowMC optimization design and validation (scope-dependent draft chapter).
\item Chapter 7 evaluates benchmark outcomes and critically reviews limitations.
\item Chapter 8 addresses legal, social, ethical, and professional issues.
\item Chapter 9 concludes and proposes future work.
\end{itemize}
