\chapter{Cipher Implementations in RISC Zero}

This chapter documents the implemented targets before optimization analysis.
The chapter is baseline-focused; final evaluation emphasis remains scope-dependent
as described in Chapter 3.
Each target follows the same high-level pattern: host code prepares a benchmark
specification, guest code executes and validates the workload, and the host
verifies the receipt and emits JSON metrics.

\section{AES Baseline (\texttt{aes-r0})}

The AES baseline target uses AES-128 test vectors and performs an
encrypt/decrypt roundtrip check in the guest. The host serializes plaintext,
key, and expected ciphertext into an \texttt{AesTestSpec} structure and validates
the guest result after receipt verification.

AES is included as a practical block-cipher comparator with well-understood
structure \cite{daemen2002rijndael}. In this repository, the AES benchmark is
currently configured with a 64-byte payload, and emits both timing and cycle
statistics.

\section{AES Optimization Workspace (\texttt{aes-r0-optimised})}

The optimized AES workspace vendors a local AES crate and introduces backend
switching for S-box experimentation in
\texttt{aesencryption/src/subbytes.rs}. Current features include
\texttt{sbox-table} (default) and \texttt{sbox-logic}.

At present, \texttt{sbox-logic} is a structural placeholder that preserves the
existing lookup-table behavior. Therefore, this branch is best interpreted as an
optimization scaffold rather than a completed optimization success.

In the current draft snapshot, baseline and optimized AES metrics are nearly identical,
which is consistent with this implementation state.

\section{Salsa20 Baseline (\texttt{salsa-r0})}

The Salsa20 target implements a manual Salsa20/20 core in
\texttt{salsa-core/src/lib.rs}, validates against a known keystream prefix, and
benchmarks encryption over a 1024-byte payload.

Salsa20 is useful here as a structural contrast to block ciphers because it is
ARX-based and avoids S-box tables \cite{bernstein2008salsa20}. This allows the
evaluation chapter to discuss not only block-vs-stream differences but also how
nonlinear realization choices appear in proving cost behavior.

\section{LowMC Baseline (\texttt{lowmc-r0})}

The baseline LowMC implementation ports reference parameters corresponding to
block size 256, key size 80, 49 S-boxes, and 12 rounds
\cite{albrecht2015ciphers}. Host code initializes plaintext/key vectors,
constructs precomputed matrices/constants from \texttt{LowMc::new}, and passes
these into the guest for encrypt/decrypt validation.

This target is especially relevant because LowMC is often discussed in contexts
where nonlinear complexity is constrained for advanced cryptographic settings.
The project benchmarks show that practical proving cost is still substantial in
this concrete implementation.

\section{Current Workload Summary}

Table \ref{tab:workloads} summarizes workload parameters as configured in the
current benchmark campaign.

\begin{table}[t]
  \begin{center}
    \begin{tabular}{|p{0.22\textwidth}|p{0.20\textwidth}|p{0.48\textwidth}|}
      \hline
      \textbf{Target} & \textbf{Algorithm} & \textbf{Configured workload parameters} \\
      \hline
      \texttt{aes-r0} & AES-128 & \texttt{payload\_bytes = 64} \\
      \hline
      \texttt{aes-r0-optimised} & AES-128 & \texttt{payload\_bytes = 64} \\
      \hline
      \texttt{salsa-r0} & Salsa20 & \texttt{payload\_bytes = 1024}, \texttt{key\_bytes = 32}, \texttt{nonce\_bytes = 8} \\
      \hline
      \texttt{lowmc-r0} & LowMC & \texttt{block\_bytes = 32}, \texttt{key\_bytes = 10} \\
      \hline
      \texttt{lowmc-optimised} & LowMC & \texttt{block\_bytes = 32}, \texttt{key\_bytes = 10} \\
      \hline
    \end{tabular}
  \end{center}
  \caption{Workload parameters in the reporting snapshot}
  \label{tab:workloads}
\end{table}

\textbf{TODO: add a host/guest/proof pipeline figure for one benchmark run,
showing where serialization, guest execution, proving, and verification occur.}
