\chapter{Project Scope, Objectives, and Success Criteria}

\section{Scope Reset from the Original Proposal}

\textbf{Working status: in progress (scope not fully locked).}

\begin{itemize}
\item Original proposal was broad (multi-zkVM + multiple cipher families + stronger ASA angle).
\item Current implemented baseline exists (RISC Zero harness + AES/Salsa20/LowMC benchmarks).
\item Active scope pivot is to narrow around LowMC optimization decisions.
\item Main candidate scope: compare LowMC styles (S-box-optimised vs more circuit-oriented).
\item Alternative candidate scope: implement one full authenticated-encryption algorithm.
\item Final report claims should only cover whichever path reaches complete implementation + evidence.
\end{itemize}

\begin{table}[t]
  \begin{center}
    \begin{tabular}{|p{0.27\textwidth}|p{0.30\textwidth}|p{0.33\textwidth}|}
      \hline
      \textbf{Area} & \textbf{Previous Implemented Baseline} & \textbf{Working Scope (Draft)} \\
      \hline
      zkVM coverage & RISC Zero only & RISC Zero only (no SP1 expansion planned) \\
      \hline
      Core technical focus & Multi-cipher benchmarking (AES/Salsa20/LowMC) & LowMC-focused optimization study \\
      \hline
      Comparison axis & Cross-cipher cost comparisons & Intra-LowMC comparison: S-box-optimised vs circuit-oriented variant \\
      \hline
      Alternative direction & Not central & Full authenticated-encryption implementation (if selected as final scope) \\
      \hline
      Reporting stance & Delivered snapshot framing & Draft notes; final claims deferred until scope lock \\
      \hline
    \end{tabular}
  \end{center}
  \caption{Scope reset and current in-progress direction}
  \label{tab:scope-reset}
\end{table}

\section{Project Objectives (Draft Notes)}

\textbf{Primary track (currently preferred): LowMC optimization comparison.}

\begin{itemize}
\item Finalize the \texttt{lowmc-sbox-optimised} implementation path and keep tests green.
\item Build/clean a more circuit-oriented LowMC implementation variant.
\item Benchmark both variants using identical harness settings and trial policy.
\item Record proving time, verification time, cycle count, and related output metrics.
\item Explain how S-box handling differences change proof-cost behavior.
\item Produce one decision note: which variant is better under this RISC Zero setup.
\end{itemize}

\textbf{Alternative track (if pivoted): full authenticated-encryption algorithm.}

\begin{itemize}
\item Implement one complete authenticated-encryption flow in guest/host path.
\item Validate correctness for encrypt + decrypt + tag verification.
\item Benchmark authenticated mode against current symmetric baseline workload.
\item Quantify overhead added by authentication/tag checks during proving.
\item Document why this track replaces or complements the LowMC variant study.
\end{itemize}

\textbf{Evidence plan (either track).}

\begin{itemize}
\item Code paths in target crates (\texttt{lowmc-*} or selected authenticated-encryption crate).
\item Harness + aggregation scripts (\texttt{bench-harness/...}).
\item Benchmark artifacts under \texttt{artifacts/benchmarks/...}.
\item Supporting notes (e.g., \texttt{notes/SBOX\_OPTIMISATIONS.md}).
\end{itemize}

\section{Non-goals and Constraints (Current)}

\begin{itemize}
\item No multi-zkVM evaluation; SP1 remains out of scope.
\item No formal ASA detection system in this phase.
\item No claim of final AES logic-backend optimization success.
\item Only one major new scope track is expected to be completed (LowMC variants or full authenticated encryption).
\item Final non-goals list must be tightened once scope lock is confirmed.
\end{itemize}

\section{Project Chronology and Pivots (Notes)}

\begin{table}[t]
  \begin{center}
    \begin{tabular}{|p{0.23\textwidth}|p{0.67\textwidth}|}
      \hline
      \textbf{Date / Window} & \textbf{Milestone Note} \\
      \hline
      2026-01-29 to 2026-01-30 & Repository initialization and early planning notes. \\
      \hline
      2026-02-26 & Monorepo consolidation and documentation restructuring. \\
      \hline
      2026-03-05 & Benchmark harness campaigns across legacy/operation targets. \\
      \hline
      2026-03-09 & Added runs including \texttt{aes-r0-optimised}, \texttt{lowmc-r0}, and \texttt{lowmc-optimised}. \\
      \hline
      2026-03-10 & Reporting snapshot \texttt{20260310-010549Z}. \\
      \hline
      Next (planned) & Scope lock: choose final track (LowMC variant comparison vs full authenticated-encryption algorithm). \\
      \hline
      Next (planned) & Run final benchmark campaign aligned to chosen track only. \\
      \hline
      Next (planned) & Freeze success criteria and remove draft-only notes. \\
      \hline
    \end{tabular}
  \end{center}
  \caption{Chronology with completed milestones and upcoming pivots}
  \label{tab:timeline}
\end{table}

\textbf{TODO: add supervisor meeting checkpoints and rationale for final scope lock decision.}

\section{Definition of Project Success (Draft)}

\textbf{Global requirements (must hold regardless of chosen track).}

\begin{enumerate}
\item reproducible benchmark workflow with stored configs, logs, and aggregates;
\item correctness tests pass for any optimized/new cryptographic implementation;
\item claims are restricted to measured repository evidence;
\item limitations and unfinished items are explicitly documented.
\end{enumerate}

\textbf{Track-specific success gate.}

\begin{itemize}
\item If LowMC track is selected: both LowMC variants are benchmarked fairly, with one defensible optimization conclusion.
\item If authenticated-encryption track is selected: complete flow is implemented, validated, and benchmarked with clear overhead analysis.
\end{itemize}
